`*--- title: "Template" author: "Studentnames and studentnumbers here" date: "`r Sys.Date()\`" output: pdf_document ---

# Set-up your environment

```{r package_install, include=FALSE}
```

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("cbsodataR")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
require(cbsodataR)
```

# Title Page

**Notitie voor de groep:** Vul hier de titel van het project, de namen van alle groepsleden, studentnummers, het tutorialgroepnummer en de naam van de tutorialdocent in. Zorg dat dit er netjes uitziet als een echte titelpagina.
```{r}
```


```{r}
```


```{r}
```


```{r}
```


```{r}
```

Include your names Andre Gouda Include the tutorial group number

Include your tutorial lecturer's name

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem

**Notitie voor de groep:** Beschrijf hier het sociale probleem dat jullie onderzoeken. Leg uit dat de woningmarkt in Nederland voor veel mensen minder betaalbaar is geworden. Bespreek waarom dit maatschappelijk relevant is, bijvoorbeeld voor starters, studenten, jonge huishoudens en middeninkomens. Leg ook kort uit waarom het nuttig is om huizenprijzen te vergelijken met besteedbaar inkomen. Gebruik hier minimaal één APA-bron om de relevantie van het probleem te ondersteunen.


The Social Problem
The Dutch housing market has become increasingly inaccessible for large groups of people in recent years. Housing is expensive for many people searching for a home  particularly for first-time buyers, people with low or middle incomes, and single-person households. House prices have risen faster than the maximum borrowing capacity based on income, which has significantly worsened housing affordability since 2013 (De Nederlandsche Bank [DNB], 2026).
The price increases are considerable: an existing owner-occupied home was on average 8.7 percent more expensive in 2024 than in 2023, and in 2025 house prices were on average a further 8.5 percent higher than the year before (Centraal Bureau voor de Statistiek [CBS], 2025a; DNB, 2026). While the average house price in 2023 was still €416,000, this increased substantially to an average of €451,000 in 2024 (BPD, n.d.).

Why is this relevant?
The affordability crisis in the housing market affects a large and diverse group of people in Dutch society. Approximately 3 million households  with an income between modal and twice the modal income  have no access to the housing market because available homes exceed their budget and there is insufficient suitable supply (BPD, n.d.).
For young people and first-time buyers, the situation is particularly acute. Buying or renting a home is becoming increasingly expensive, making it difficult for first-time buyers to find housing, as reflected in the growing share of young adults between the ages of 25 and 30 still living with their parents (CBS, 2025b). The difficulties that young people and first-time buyers currently face in entering the housing market are clearly visible in homeownership statistics: it is not until the 35 to 39 age group that the same level of homeownership as in previous years is seen again (ABF Research, 2025).
Students and young households are also severely affected by the tightness in the rental market. The number of rental transactions in the private sector was as much as 38 percent lower at the end of 2025 than a year earlier, and in the mid-range rental sector the number of homes let in the four major cities fell by as much as 50 to 70 percent (NVM, 2026).

Why compare house prices with disposable income?
In order to truly understand the affordability of the housing market, it is necessary to relate house prices to the disposable income of households. House prices are rising faster than incomes, which reduces the purchasing power of home buyers (Garantiemakelaars, 2025). Only by placing these two figures side by side does it become clear how large the gap truly is  and for whom housing risks becoming unattainable. An absolute price of €451,000 says little without knowing what a household can put against it each month.

Include the following:

- Why is this relevant?

- ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

**Notitie voor de groep:** Leg hier uit welke datasets jullie gebruiken en waar ze vandaan komen. Noem dat de data afkomstig is van CBS StatLine. Beschrijf kort dat één dataset informatie geeft over gemiddelde verkoopprijzen van bestaande koopwoningen en dat de andere dataset informatie geeft over gemiddeld besteedbaar inkomen van huishoudens. Vermeld ook waarom deze twee datasets samen nodig zijn voor jullie onderzoek.

Which datasets do we use and where do they come from?
For this research, we make use of two datasets that both originate from CBS StatLine, the official statistics database of Statistics Netherlands (Centraal Bureau voor de Statistiek [CBS], n.d.a). CBS StatLine is a reliable and widely recognised source within the Netherlands, which is regularly updated with current figures on a wide range of social and economic topics.

Dataset 1: Average selling prices of existing owner-occupied homes
The first dataset contains information on the average selling prices of existing owner-occupied homes in the Netherlands. This dataset provides insight into how house prices have developed over the years, broken down by region and housing type (CBS, n.d.a). This allows us to map the price developments on the Dutch housing market.
Dataset 2: Average disposable income of households
The second dataset contains information on the average disposable income of households in the Netherlands. Disposable income refers to the amount of money households actually have available after the deduction of taxes and social contributions (CBS, n.d.b). This dataset makes it possible to map the financial position of different household groups.

Why are both datasets needed together?
Both datasets are necessary in order to answer the central question of our research. A house price on its own says little about the affordability of the housing market. Only when the development of house prices is compared to the development of disposable income does it become clear whether housing has actually become less affordable for households (De Nederlandsche Bank [DNB], 2026). By combining the two datasets, we can calculate to what extent the growth in house prices has outpaced the growth in incomes, and what this means for groups such as first-time buyers, students, young households, and middle-income earners (CBS, n.d.a; CBS, n.d.b).

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}
dataset1 <- cbs_get_data("83625NED")
dataset2 <- cbs_get_data("86161NED")
```
 
 
 
```{r}
metadata1 <- cbs_get_meta("83625NED")
metadata2 <- cbs_get_meta("86161NED")
populatie_meta <- metadata2$Populatie
regios_meta <- metadata2$RegioS
huishouden_meta <- metadata2$KenmerkenVanHuishoudens
```


```{r}
names(metadata1)
names(populatie_meta)
```


```{r}
names(metadata2)
names(regios_meta)
names(huishouden_meta)
```

```{r}
populatie_meta %>%
  filter(Key %in% unique(dataset2$Populatie))
```


```{r}
huishouden_meta %>%
  filter(Key %in% unique(dataset2$KenmerkenVanHuishoudens))
```

## 2.2 Provide a short summary of the dataset(s)

**Notitie voor de groep:** Geef hier een korte samenvatting van beide datasets. Beschrijf hoeveel rijen en kolommen de datasets ongeveer hebben, welke jaren erin zitten, welke regio’s worden gebruikt en welke variabelen belangrijk zijn voor jullie analyse. Leg vooral de betekenis uit van `RegioS`, `Perioden`, `GemiddeldeVerkoopprijs_1` en `GemiddeldBesteedbaarInkomen_5`. Vermeld ook de meeteenheden, bijvoorbeeld euro’s en duizend euro’s.
Dataset 1: Average selling prices of existing owner-occupied homes
The first dataset (dataset1) contains data on the average selling prices of existing owner-occupied homes in the Netherlands (Centraal Bureau voor de Statistiek [CBS], n.d.a). The dataset consists of 23,095 rows and 3 columns. The years in the dataset range from 1995 to the most recently available years. The most important variables are:

RegioS: this variable indicates the region to which the data applies. In this dataset, the code "NL01" is used, which represents the Netherlands as a whole.
Perioden: this variable indicates the year over which the measurement was taken, displayed in the format "1995JJ00", "1996JJ00", and so on. "JJ00" refers to an annual average.
GemiddeldeVerkoopprijs_1: this is the most important variable for our research. It represents the average selling price of existing owner-occupied homes in the Netherlands, expressed in euros. In 1995, for example, this price was €93,750, rising to €172,050 in the year 2000 (CBS, n.d.a).

Dataset 2: Average disposable income of households
The second dataset (dataset2) contains data on the income of households in the Netherlands (CBS, n.d.b). The dataset consists of 6 rows and 20 columns. Based on the visible observations, the years in the dataset range from 2011 to 2016. The most important variables are:

Populatie: a code indicating the population group to which the measurement relates.
KenmerkenVanHuishoudens: a code describing the type of household, for example single-person households or multi-person households.
RegioS: just as in dataset 1, this variable indicates the region. Here too, "NL01" is used for the Netherlands as a whole (CBS, n.d.b).
Perioden: the measurement year, displayed in the same format as dataset 1, for example "2011JJ00".
GemiddeldBesteedbaarInkomen_5: this is the most important variable for our research. It represents the average disposable income of households, expressed in thousands of euros. For example, the average disposable income in 2011 was approximately €36,600 (displayed as 36.6 in the dataset) (CBS, n.d.b).

Why are these variables relevant together?
By linking the variable GemiddeldeVerkoopprijs_1 from dataset 1 to the variable GemiddeldBesteedbaarInkomen_5 from dataset 2, through the shared variables RegioS and Perioden, we can analyse per year how the ratio between house prices and disposable income has developed (De Nederlandsche Bank [DNB], 2026). It is important to note the difference in units of measurement: the selling prices are expressed in full euros, while the disposable income is expressed in thousands of euros. This difference must be corrected for during the analysis (CBS, n.d.a; CBS, n.d.b).
```{r}
head(dataset1)

names(dataset1)

glimpse(dataset1)

```

<<<<<<< HEAD
=======

```{r}
head(dataset2)

names(dataset2)

glimpse(dataset2)
```

>>>>>>> 45f98fcac9f46a0a08702354b3dc3f499f29d04d
**Notitie voor de groep:** Gebruik de output van `head()`, `names()` en `glimpse()` om in gewone zinnen te beschrijven hoe de data eruitziet. Schrijf niet alleen dat er kolommen zijn, maar leg uit welke kolommen jullie echt gebruiken en waarom.

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

**Notitie voor de groep:** Beschrijf hier het type variabelen. `RegioS` is een categorische variabele, `Perioden` is een tijdsvariabele, en de huizenprijs en het besteedbaar inkomen zijn numerieke variabelen. Leg ook uit dat de data geaggregeerde regionale data is en dus geen persoonlijke gegevens van individuele mensen bevat. Benoem dat inkomen kan worden gezien als een sociaaleconomische variabele.

Think of things like:

- Do the variables contain health information or SES information?

- Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

The dataset contains a mix of variable types. "RegioS" is a categorical variable that indicates the region, and "Perioden" is a time variable that shows the year of observation. Both "GemiddeldeVerkoopprijs_1" and "GemiddeldBesteedbaarInkomen_5" are numerical variables.
The housing price data comes from administrative sources, specifically transaction records from the Kadaster (Land Registry). This means the prices reflect actual sales, not asking prices. The disposable income variable is a socioeconomic variable, also from administrative sources like tax records, and is measured in thousands of euros per household per year.
One limitation worth noting is that both variables are measured at the regional level, so they represent averages. This means we cannot say anything about individual households. A region with a high average income may still have many people who struggle to afford a house there.


# Part 3 - Quantifying

## 3.1 Data cleaning

**Notitie voor de groep:** Leg hier uit welke stappen jullie nemen om de data schoon te maken. Beschrijf dat jullie alleen de relevante kolommen selecteren, missende waarden verwijderen en de twee datasets samenvoegen op regio en periode. Leg ook uit waarom jullie filteren op specifieke huishoudcategorieën in de inkomensdataset. Noem dat `inner_join()` alleen observaties bewaart die in beide datasets voorkomen.

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

Before merging the two datasets, we first selected only the columns we actually need. For the housing dataset we kept "RegioS", "Perioden" and "GemiddeldeVerkoopprijs_1", and for the income dataset we kept "RegioS", "Perioden" and "GemiddeldBesteedbaarInkomen_5".
We also filtered the income dataset on a specific household type, because the original dataset contains multiple population categories. We only kept the category that represents all private households, to make sure we are comparing the right group to the housing prices.
After selecting the relevant columns, we removed all rows with missing values in both datasets using "drop_na()". This is important because missing values in either the price or income variable would make the analysis unreliable. After cleaning, the housing dataset had "r nrow(housing)" rows and the income dataset had "r nrow(income)" rows.
Finally, we merged the two datasets using "inner_join()" on both "RegioS" and "Perioden". This means we only kept observations that appear in both datasets. The combined dataset ended up with "r nrow(combined)" rows, which means some regions or years were lost because they did not appear in both datasets. This is a limitation we should keep in mind when interpreting the results.



```{r data_cleaning}
housing <- dataset1 %>%
  select(
    RegioS,
    Perioden,
    GemiddeldeVerkoopprijs_1
  )

head(housing)
```


```{r data_cleaning}
income <- dataset2 %>%
  filter(
    Populatie == "1050010",
    KenmerkenVanHuishoudens == "1050010"
  ) %>%
  select(
    RegioS,
    Perioden,
    GemiddeldBesteedbaarInkomen_5
  ) 

head(income)
```


```{r data_cleaning}
housing <- housing %>%
  drop_na()

income <- income %>%
  drop_na()
```


```{r data_cleaning}
nrow(housing)
unique(housing$RegioS)[1:20]
```


```{r data_cleaning}
nrow(income)
unique(income$RegioS)[1:20]
```


```{r data_cleaning}
combined <- housing %>%
  inner_join(
    income,
    by = c("RegioS", "Perioden")
  )
head(combined)
```


```{r data_cleaning}
summary(combined)
```


```{r data_cleaning}
unique(combined$Perioden)
```


```{r data_cleaning}
unique(dataset2$Populatie)
```


```{r data_cleaning}
unique(dataset2$KenmerkenVanHuishoudens)
unique(dataset2$Populatie)

```

**Notitie voor de groep:** Schrijf onder deze code kort wat er is gebeurd. Bijvoorbeeld: eerst is een aparte housing-dataset gemaakt, daarna een aparte income-dataset. Vervolgens zijn missende waarden verwijderd en zijn de datasets samengevoegd. Gebruik de output van `nrow()` om te beschrijven hoeveel observaties overblijven na het opschonen.

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

<<<<<<< HEAD
Variable 1 - Year
**Notitie voor de groep:** Leg hier uit waarom jullie de variabele `Year` maken. De originele CBS-periode staat bijvoorbeeld als `2011JJ00`, maar voor grafieken is een numeriek jaar makkelijker te gebruiken. Beschrijf dat jullie met `substr()` de eerste vier tekens uit `Perioden` halen.
=======
Variable 1 **Notitie voor de groep:** Leg hier uit waarom jullie de variabele `Year` maken. De originele CBS-periode staat bijvoorbeeld als `2011JJ00`, maar voor grafieken is een numeriek jaar makkelijker te gebruiken. Beschrijf dat jullie met `substr()` de eerste vier tekens uit `Perioden` halen.
>>>>>>> fa33898693ec57884c04626aa23c73a881a19eaf


The first variable we created is "Year". In the original CBS data, the time variable "Perioden" is stored as a character string in the format `2011JJ00`, which is not directly usable in plots or calculations. By extracting the first four characters using "substr()" and converting them to a number with "as.numeric()", we get a clean numeric year variable that we can use on the x-axis of our time series plots.


```{r gen_var1}
combined <- combined %>%
  mutate(
    Year = as.numeric(substr(Perioden, 1, 4))
  )

head(combined)
```

<<<<<<< HEAD
Variable 2 - AffordabilityRatio
**Notitie voor de groep:** Leg hier uit waarom jullie de `AffordabilityRatio` maken. Deze ratio vergelijkt de gemiddelde huizenprijs met het gemiddelde besteedbaar inkomen. Een hogere ratio betekent dat woningen minder betaalbaar zijn. Leg ook uit waarom het inkomen met 1000 wordt vermenigvuldigd: de inkomensvariabele staat in duizenden euro’s.
=======
Variable 2 **Notitie voor de groep:** Leg hier uit waarom jullie de `AffordabilityRatio` maken. Deze ratio vergelijkt de gemiddelde huizenprijs met het gemiddelde besteedbaar inkomen. Een hogere ratio betekent dat woningen minder betaalbaar zijn. Leg ook uit waarom het inkomen met 1000 wordt vermenigvuldigd: de inkomensvariabele staat in duizenden euro’s.
>>>>>>> fa33898693ec57884c04626aa23c73a881a19eaf

The second variable is "AffordabilityRatio". This variable divides the average house price by the average disposable income per household. Because the income variable is stored in thousands of euros, we multiply it by 1000 before dividing. The result tells us how many annual incomes a household would need to buy an average house in that region and year. A higher ratio means housing is less affordable. This variable is central to our analysis because it combines both datasets into one meaningful measure of the housing problem.
For example, a ratio of 8 means that an average household would need 8 years of their entire disposable income to buy an average house, without spending anything else. This makes the affordability problem more concrete.


```{r gen_var2}
combined <- combined %>%
  mutate(
    AffordabilityRatio = GemiddeldeVerkoopprijs_1 / (GemiddeldBesteedbaarInkomen_5 * 1000)
  )
head(combined)
```


```{r gen_var2}
summary(combined$AffordabilityRatio)
```

## 3.3 Visualize temporal variation
<<<<<<< HEAD

**Notitie voor de groep:** Beschrijf hier wat de tijdsgrafieken laten zien. De eerste grafiek laat zien hoe de gemiddelde huizenprijzen zich door de tijd ontwikkelen. De tweede grafiek laat zien hoe de betaalbaarheid verandert door de tijd. Bespreek of huizenprijzen stijgen of dalen en of woningen relatief duurder worden ten opzichte van inkomen. Koppel dit terug aan de woningcrisis.
=======

The first graph shows how average house prices in the Netherlands changed over time. Between 2011 and 2013, house prices went down from about €250,000 to €220,000. After 2014, prices started to rise again. The increase became much stronger after 2017, and between 2020 and 2022 prices grew very quickly. Although there was a small drop in 2023, prices increased again in 2024 to around €450,000, the highest point in the graph.

The second graph shows how housing affordability changed over time. If affordability goes down, it means that houses become more expensive compared to people's incomes. Even if incomes increase, house prices may rise faster, making it harder for people to buy a home.

Together, the graphs show that house prices have increased much more than incomes. As a result, people need to spend a larger part of their income on housing or take out bigger mortgages. This is especially difficult for first-time buyers and people with average incomes.

These trends are closely related to the housing crisis in the Netherlands. Because there are not enough homes available, demand is higher than supply. This pushes house prices up and makes housing less affordable. The housing crisis can be seen in high house prices, long waiting lists for rental homes, and the difficulties many people face when trying to buy a house. Overall, the graphs show that housing has become less affordable over the years.

```{r}
yearly_prices <- combined %>%
  group_by(Year) %>%
  summarise(
    AveragePrice = mean(GemiddeldeVerkoopprijs_1, na.rm = TRUE)
  )
```


```{r}
ggplot(yearly_prices, aes(x = Year, y = AveragePrice)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(
    breaks = yearly_prices$Year
  ) +
  labs(
    title = "Average House Prices in the Netherlands (2011-2024)",
    x = "Year",
    y = "Average House Price (€)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```




## 3.4 Visualize spatial variation

**Notitie voor de groep:** Leg hier uit waarom regionale verschillen belangrijk zijn. Deze grafiek laat zien welke regio’s in 2024 de hoogste gemiddelde huizenprijzen hebben. Bespreek of sommige regio’s duidelijk duurder zijn dan andere en waarom dit relevant is voor het sociale probleem van woningbetaalbaarheid.
```{r}
regio_names <- regios_meta %>%
  select(
    RegioS = Key,
    RegionName = Title
  ) %>%
  mutate(
    RegioS = trimws(RegioS)
  )

combined <- combined %>%
  mutate(
    RegioS = trimws(RegioS)
  ) %>%
  left_join(
    regio_names,
    by = "RegioS"
  )
```

```{r visualise_map}
top_regions <- combined %>%
  filter(Year == 2024) %>%
  arrange(desc(GemiddeldeVerkoopprijs_1)) %>%
  slice(1:15)
```


```{r visualise_map}
ggplot(top_regions,
       aes(
         x = reorder(RegionName, GemiddeldeVerkoopprijs_1),
         y = GemiddeldeVerkoopprijs_1
       )) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 15 Regions by Average House Price in 2024",
    x = "Region",
    y = "Average House Price (€)"
  ) +
  theme_minimal()
```

=======
**Notitie voor de groep:** Schrijf hier in eigen woorden wat de grafiek laat zien. Noem de belangrijkste regio’s uit de grafiek en leg uit waarom regionale verschillen in huizenprijzen belangrijk zijn voor jullie onderzoek.
>>>>>>> 84138f02eecc55c61867d316f1cce39f0461d2d1
**Notitie voor de groep:** Schrijf hier in eigen woorden wat de grafiek laat zien. Noem de belangrijkste regio’s uit de grafiek en leg uit waarom regionale verschillen in huizenprijzen belangrijk zijn voor jullie onderzoek.
Here you provide a description of why the plot above is relevant to your specific social problem.


```{r}
gemeente_map <- cbs_get_sf(
  region = "gemeente",
  year = 2024
)

map_data <- combined %>%
  filter(Year == 2024)

map_combined <- gemeente_map %>%
  left_join(
    map_data,
    by = c("statcode" = "RegioS")
  )

top15_codes <- top_regions$RegioS

map_combined <- map_combined %>%
  mutate(
    Top15 = ifelse(statcode %in% top15_codes,
                   "Top 15",
                   "Other municipalities")
  )
ggplot(map_combined) +
  geom_sf(aes(fill = Top15),
          color = "white",
          linewidth = 0.05) +
  scale_fill_manual(
    values = c(
      "Top 15" = "red",
      "Other municipalities" = "grey"
    )
  ) +
  labs(
    title = "Top 15 Most Expensive Municipalities in 2024",
    fill = ""
  ) +
  theme_minimal()
```

## 3.5 Visualize sub-population variation

What is the poverty rate by state? **Notitie voor de groep:** Pas deze tekst aan, want jullie onderzoeken geen poverty rate by state. Jullie kunnen hier uitleggen hoe besteedbaar inkomen verschilt tussen regio’s of inkomensgroepen. Beschrijf waarom inkomensverschillen belangrijk zijn voor woningbetaalbaarheid: een regio met hoge huizenprijzen is vooral problematisch als het inkomen daar niet even sterk meegroeit.


```{r}
income_2024 <- combined %>%
  filter(Year == 2024) %>%
  mutate(
    PriceGroup = ntile(GemiddeldeVerkoopprijs_1, 5)
  )
```

```{r}
ggplot(
  income_2024,
  aes(
    x = factor(PriceGroup),
    y = GemiddeldBesteedbaarInkomen_5
  )
) +
  geom_boxplot() +
  labs(
    title = "Disposable Income by House Price Group (2024)",
    x = "House Price Group",
    y = "Disposable Income (thousand €)"
  ) +
  theme_minimal()
```


Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

**Notitie voor de groep:** Deze sectie is eigenlijk een relatie-analyse tussen twee variabelen. Leg uit dat jullie de relatie onderzoeken tussen gemiddeld besteedbaar inkomen en gemiddelde huizenprijs. Bespreek of regio’s met hogere inkomens ook hogere huizenprijzen hebben. Leg ook uit dat een positief verband niet automatisch betekent dat woningen betaalbaarder zijn.

Analyze the relationship between two variables. 

```{r}
yearly_affordability <- combined %>%
  group_by(Year) %>%
  summarise(
    AverageAffordability = mean(AffordabilityRatio)
  )
```


```{r}
ggplot(yearly_affordability,
       aes(x = Year,
           y = AverageAffordability)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(
    breaks = yearly_affordability$Year
  ) +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "red") +
  annotate("text",
           x = 2020, y = max(yearly_affordability$AverageAffordability, na.rm = TRUE) * 0.97, label = "COVID period", angle = 90, vjust = -0.5, size = 3) +
  labs(
    title = "Average Housing Affordability Ratio (2011-2024)",
    x = "Year",
    y = "House Price / Disposable Income"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

**Notitie voor de groep:** Vat hier de belangrijkste resultaten samen. Bespreek wat jullie grafieken laten zien over de ontwikkeling van huizenprijzen, regionale verschillen en de verhouding tussen huizenprijzen en inkomen. Beantwoord de centrale vraag: zijn woningen minder betaalbaar geworden of zijn er duidelijke regionale verschillen? Bespreek ook beperkingen van jullie analyse, bijvoorbeeld dat de data geaggregeerd is en geen individuele huishoudens volgt.

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ... **Notitie voor de groep:** Plaats hier de link naar jullie publieke GitHub repository. Controleer dat de repository openbaar is en dat iemand anders de code kan openen. Vermeld kort dat de R Markdown file, code en instructies daar beschikbaar zijn.

## 5.2 Reference list

**Notitie voor de groep:** Zet hier alle bronnen in APA-stijl. Voeg in ieder geval de twee CBS StatLine datasets toe die jullie gebruiken. Voeg ook bronnen toe die jullie gebruiken om de woningcrisis of betaalbaarheid van woningen te onderbouwen. Zorg dat alle bronnen die in de tekst worden genoemd ook in de literatuurlijst staan.

Use APA referencing throughout your document.

Use APA referencing throughout your document.

ABF Research & Ministerie van Binnenlandse Zaken. (2025). Woningmarktverkenning 2025–2040. https://open.overheid.nl/documenten/a639725d-2ae1-4fef-b4ee-0023c6b62f4a/file
BPD. (z.d.). Betaalbaar wonen. https://www.bpd.nl/betaalbaar
Centraal Bureau voor de Statistiek. (2025a). Bestaande koopwoningen bijna 9 procent duurder in 2024. https://www.cbs.nl/nl-nl/nieuws/2025/04/bestaande-koopwoningen-bijna-9-procent-duurder-in-2024
Centraal Bureau voor de Statistiek. (2025b). Wonen: Monitor Brede Welvaart en de Sustainable Development Goals. https://www.cbs.nl/nl-nl/visualisaties/monitor-brede-welvaart-en-de-sustainable-development-goals/hier-en-nu/wonen
De Nederlandsche Bank. (2026). De woningmarkt. https://www.dnb.nl/actuele-economische-vraagstukken/woningmarkt/
Garantiemakelaars. (2025, 10 februari). Terugblik en vooruitblik: Wat brengt de woningmarkt in 2025? https://garantiemakelaars.nl/wat-brengt-2025/
NVM. (2026, maart). Uitponden vergroot ongelijkheid starters: huurwoning verdwijnt, koopwoning onbereikbaar. https://www.nvm.nl/nieuws/2026/nvm-uitponden-vergroot-ongelijkheid-starters-huurwoning-verdwijnt-koopwoning-onbereikbaar/

Centraal Bureau voor de Statistiek. (z.d.). CBS StatLine [Dataset]. https://opendata.cbs.nl/statline/
Centraal Bureau voor de Statistiek. (z.d.a). Bestaande koopwoningen; verkoopprijzen prijsindex [Dataset]. CBS StatLine. https://opendata.cbs.nl/statline/#/CBS/nl/dataset/83906NED/table
Centraal Bureau voor de Statistiek. (z.d.b). Inkomen van huishoudens; besteedbaar inkomen [Dataset]. CBS StatLine. https://opendata.cbs.nl/statline/#/CBS/nl/dataset/84959NED/table
De Nederlandsche Bank. (2026). De woningmarkt. https://www.dnb.nl/actuele-economische-vraagstukken/woningmarkt/